IIT CS595: Topics \& Applications in Programming Languages\\ {\large Homework 0: Rule induction, expressions, type safety}

### 2021-09-17 HW0/hw0.tex
fix prof typo, task 3
HW0 done

Inductive case assuming $\nodes{\tree} \geq 2^{n-1}$:
\begin{itemize}
    \item Case when tree's top element is a black node
        \[
        \begin{array}{l l l l}
            \nodes{\bnode{\tree_1}{\tree_2}} & \geq & 2^{\kwlen{\tree}-1} & \text{the proposition for this case}\\
            1 + \nodes{\tree_1} + \nodes{\tree_2} & \geq & 1 + 2 ^ {\kwlen{\tree_1} - 1} + 2 ^ {\kwlen{\tree_2} - 1} & \text{apply rule (RBT-2)}\\
            1 & \geq & 1  & \text{apply inductive hypothesis}\\
        \end{array}
        \]

    \item Case when tree's top element is a red node
        \[
        \begin{array}{l l l l}
            \nodes{\rnode{\tree_1}{\tree_2}} & \geq & 2^{\kwlen{\tree}-1} & \text{the proposition for this case}\\
            1 + \nodes{\tree_1} + \nodes{\tree_2} & \geq & 2 ^ {\kwlen{\tree_1} - 1} + 2 ^ {\kwlen{\tree_2} - 1} & \text{apply rule (RBT-3)}\\
            1 & \geq & 0  & \text{apply inductive hypothesis}\\
        \end{array}
        \]
\end{itemize}

@@

@@

{
  \centering
  \def \MathparLineskip {\lineskip=0.43cm}
  \begin{mathpar}
    \Rule{T-8}
         {\typed{\ctx}{e_1}{\kwint}\\
           % Use \\ to separate multiple premises
           \typed{\ctx}{e_2}{\kwint}}
         {\typed{\ctx}{e_1 = e_2}{\kwbool}}
    \and
    \Rule{T-9}
         {\strut}
         {\typed{\ctx}{\kwtrue}{\kwbool}}
    \and
    \Rule{T-10}
         {\strut}
         {\typed{\ctx}{\kwtrue}{\kwbool}}
    \and
    \Rule{T-11}
         {\typed{\ctx}{e_1}{\kwbool}\\
           \typed{\ctx}{e_2}{\tau}\\
           \typed{\ctx}{e_3}{\tau}}
         {\typed{\ctx}{\kwif{e_1}{e_2}{e_3}}{\tau}}
  \end{mathpar}
}

@@

@@

@@

>>>>

Inductive case assuming $\nodes{\tree} \geq 2^{n-1}$:
\begin{itemize}
    \item Case when $\tree = \bnode{...}{...}$ -- tree's top element is a black node
        \[
        \begin{array}{l l l l}
            \nodes{\bnode{\tree_1}{\tree_2}} & \geq & 2^{\kwlen{\tree}-1} & \text{the proposition for this case}\\
            1 + \nodes{\tree_1} + \nodes{\tree_2} & \geq & 1 + 2 ^ {\kwlen{\tree_1} - 1} + 2 ^ {\kwlen{\tree_2} - 1} & \text{apply rule (RBT-2)}\\
            1 & \geq & 1  & \text{apply inductive hypothesis}\\
        \end{array}
        \]

    \item Case when $\tree = \rnode{...}{...}$ -- tree's top element is a red node
        \[
        \begin{array}{l l l l}
            \nodes{\rnode{\tree_1}{\tree_2}} & \geq & 2^{\kwlen{\tree}-1} & \text{the proposition for this case}\\
            1 + \nodes{\tree_1} + \nodes{\tree_2} & \geq & 2 ^ {\kwlen{\tree_1} - 1} + 2 ^ {\kwlen{\tree_2} - 1} & \text{apply rule (RBT-3)}\\
            1 & \geq & 0  & \text{apply inductive hypothesis}\\
        \end{array}
        \]
\end{itemize}

@@

{
    \centering
    \def \MathparLineskip{\lineskip=0.43cm}
    \begin{mathpar}
        \Rule{V-3}
            {\strut}
            {\kwfalse \val}
        \and
        \Rule{V-4}
            {\strut}
            {\kwtrue \val}
        \and
        \Rule{S-15}
            {e_1 \step e_1'}
            {\kwif{e_0}{e_1}{e_2} \step \kwif{e_0'}{e_1}{e_2}}
        \and
        \Rule{S-16}
            {\strut}
            {\kwif{\kwtrue}{e_1}{e_2} \step e_1}
        \and
        \Rule{S-17}
            {\strut}
            {\kwif{\kwfalse}{e_1}{e_2} \step e_2}
    \end{mathpar}
}

@@

{
  \centering
  \def \MathparLineskip {\lineskip=0.43cm}
  \begin{mathpar}
    \Rule{T-8}
         {\typed{\ctx}{e_1}{\kwint}\\
           % Use \\ to separate multiple premises
           \typed{\ctx}{e_2}{\kwint}}
         {\typed{\ctx}{e_1 = e_2}{\kwbool}}
    \and
    \Rule{T-9}
         {\strut}
         {\typed{\ctx}{\kwtrue}{\kwbool}}
    \and
    \Rule{T-10}
         {\strut}
         {\typed{\ctx}{\kwfalse}{\kwbool}}
    \and
    \Rule{T-11}
         {\typed{\ctx}{e_1}{\kwbool}\\
           \typed{\ctx}{e_2}{\tau}\\
           \typed{\ctx}{e_3}{\tau}}
         {\typed{\ctx}{\kwif{e_1}{e_2}{e_3}}{\tau}}
  \end{mathpar}
}

@@

Proposition: if $\typed{\ctx}{e}{\tau}$ and $e \step e'$ then $\typed{\ctx}{e'}{\tau}$
\begin{itemize}
    \item S-16: When $e = \kwif{\kwtrue}{e_1}{e_2}$\\ \step $e' = e_1$\\
        by inversion on rule (T-11) $\typed{\ctx}{\kwif{e_0}{e_1}{e_2}}{\tau}$ 
            and  $\typed{\ctx}{e_1}{\tau}$\\
        with induction on $e_1 \step e_1'$ this confirms that the type of the stepped value is preserved.

    \item S-17: When $e = \kwif{\kwfalse}{e_1}{e_2}$\\ \step $e' = e_2$\\
        by inversion on rule (T-11) $\typed{\ctx}{\kwif{e_0}{e_1}{e_2}}{\tau}$ 
            and  $\typed{\ctx}{e_2}{\tau}$\\
        with induction on $e_2 \step e_2'$ this confirms that the type of the stepped value is preserved.

    \item S-15: When $e = \kwif{e_0}{e_1}{e_2}$\\ \step $e' = \kwif{e_0'}{e_1}{e_2}$\\
        by inversion on rule (T-11), $\typed{\ctx}{e_0}{\kwbool}$, $\typed{\ctx}{e_1}{\tau}$, 
            and $\typed{\ctx}{e_2}{\tau}$\\
        by induction we can assume $e_0 \step e_0'$ with $\typed{\ctx}{e_0'}{\kwbool}$\\
        thus we can apply (T-11) to assert that the type is the same.
\end{itemize}

@@

\textbf{Answer:}
if $e \val$ and $\typed{}{e}{\kwbool}$ then $e$ is either $\kwtrue$ or $\kwfalse$.

@@

Proposition: if $\typed{}{e}{\tau}$ then $e \val$ or $e \step e'$
\begin{itemize}
    \item T-9: $e = \kwtrue$, $\tau = \kwbool$\\
        by V-4 $e \val$
    \item T-10: $e = \kwfalse$, $\tau = \kwbool$\\
        by V-3 $e \val$
    \item T-11: $e = \kwif{e_1}{e_2}{e_3}$\\
        by inversion on rule T-11 $\typed{}{e_1}{\kwbool}$\\
        by induction, either $e_1 \val$ or $e_1 \step e_1'$\\
        by the Canonical Forms lemma for from Task~\ref{task-cf} $e_1$ can hold either \kwtrue or \kwfalse
        \begin{itemize}
            \item \kwtrue : S-16 confirms that $\kwif{\kwtrue}{e_2}{e_3} \step e_2$ and thus $e' = e_2$
            \item \kwfalse : S-17 confirms that $\kwif{\kwfalse}{e_2}{e_3} \step e_3$ and thus $e' = e_3$
        \end{itemize}

\end{itemize}
